---
title: "Germline variant analysis via resampling"
output: html_document
date: "2025-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following documentation shows the methodology for the meta-analysis of germline variants in 
acute leukemia and myelodysplastic syndromes as described in Harmon, et. al, 2025:
"Germline Variant Burden Warrants Universal Genetic Testing in Pediatric Myeloid Leukemia".

The purpose of this analysis is to determine if there are significant differences
in germline predisposition between acute lymphoblastic leukemia (ALL), acute myeloid leukemia (AML), 
and myelodysplastic syndrome (MDS). To investigate this, we compared the percentage of 
patients that have a pathogenic germline variant related to hematopoietic malignancy
predisposition within published cohorts of patients.

We collected and harmonized several large WGS/WES studies to investigate
evidence for disparities in genetic variants between acute lymphoblastic leukemia,
acute myeloid leukemia, and myelodysplastic syndrome. 
Here, we focus on germline pathogenic/likely pathogenic (P/LP) variants
with seemingly relevant consequences for blood and immune development based on
the University of Chicago's genetic services panel for hereditary hematopoietic
malignancy.

This document shows the statistical methodology employed, including Poisson regression
and resampling. 

## Modeling germline burden with Poisson regression

We used Poisson regression to model germline variants because Poisson models are 
useful for modeling rare events. We used the robust Poisson test for trend, because
it accounts for differences in dispersion between the groups. We tested for a 
significance in trend (B-ALL < T-ALL < AML < MDS).

```{r, warning=FALSE, message=FALSE}
combined <- read.csv("pan_acute_leukemia_MDS_G_PLP.csv")
combined$Source <- relevel(relevel(relevel(factor(combined$Source), 4), 4), 4)
combined$Lineage <- relevel(relevel(factor(combined$Lineage), 3), 3)
combined$Group <- paste(combined$AgeGroup, gsub("\\-", "", combined$Lineage))
combined$Group <- relevel(relevel(factor(combined$Group), 4), 4)
with(combined, table(Group))

#Calculate actual percents and confidence intervalsof patients with a germline variant for each diagnosis
nrow(combined[combined$Lineage=="AML" & combined$G_PLP != 0,]) / nrow(combined[combined$Lineage=="AML",])*100
binom.test(nrow(combined[combined$Lineage=="AML" & combined$G_PLP != 0,]) , nrow(combined[combined$Lineage=="AML",]), conf.level = 0.95)

nrow(combined[combined$Lineage=="T-ALL" & combined$G_PLP != 0,]) / nrow(combined[combined$Lineage=="T-ALL",])*100
binom.test(nrow(combined[combined$Lineage=="T-ALL" & combined$G_PLP != 0,]), nrow(combined[combined$Lineage=="T-ALL",]), conf.level = 0.95)

nrow(combined[combined$Lineage=="B-ALL" & combined$G_PLP != 0,]) / nrow(combined[combined$Lineage=="B-ALL",])*100
binom.test(nrow(combined[combined$Lineage=="B-ALL" & combined$G_PLP != 0,]), nrow(combined[combined$Lineage=="B-ALL",]), conf.level = 0.95)

nrow(combined[combined$Lineage=="MDS" & combined$G_PLP != 0,]) / nrow(combined[combined$Lineage=="MDS",]) * 100
binom.test(nrow(combined[combined$Lineage=="MDS" & combined$G_PLP != 0,]), nrow(combined[combined$Lineage=="MDS",]), conf.level = 0.95)

# relevel to compare against B-ALL (largest group and lowest mutation rate)
combined$trend <- ordered(combined$Lineage)

library(robust)
fit_rob_trend <- glmRob(G_PLP ~ trend, 
                        data=subset(combined, AgeGroup == "Pediatric"),
                        family="poisson")
summary(fit_rob_trend) 

fit_rob_null <- glmRob(G_PLP ~ 1, 
                       data=subset(combined, AgeGroup == "Pediatric"),
                       family="poisson")
anova(fit_rob_trend, fit_rob_null, test = "Chisq")

```

The robust Poisson model shows a p-value < 0.001. We can therefore reject the null
hypothesis that there is no relationship between type of malignancy and frequency
of germline variants. This supports the conclusion that the frequency of germline 
variants in AML is higher than that of B-ALL and T-ALL, and lower than that of MDS.

## Resampling-based inference of germline burden

We also approached this using an orthogonal brute-force non-parametric 
approach, which is to resample over and over without replacement. The advantage 
of this approach is that resampling makes no assumptions about the underlying 
distribution of the data. 

This tests the hypothesis that the percentage of subjects with at least one
germline variant is different between ALL, AML, and MDS.

```{r, warning=FALSE, message=FALSE}
combined <- read.csv("pan_acute_leukemia_MDS_G_PLP.csv")
combined$Group <- paste(combined$AgeGroup, gsub("\\-", "", combined$Lineage))
table(combined$Group)
#Not enough adult ALL to resample
combined <- combined[!combined$Group %in% c("Adult TALL","Adult BALL"),]
combined$Group <- factor(combined$Group, levels=c("Pediatric BALL","Pediatric TALL","Pediatric AML","Adult AML","Pediatric MDS","Adult MDS"))
byGroup <- split(combined$G_PLP, combined$Group)

rs <- function(y, n, m) replicate(m, sum(y[sample.int(n=length(y),size=n)]>0)/n)

library(reshape2)
results <- melt(data.frame(lapply(byGroup, rs, n=200, m=4000)), 
                variable.name="Group", value.name="Burden")

library(ggplot2)
library(ggbeeswarm)
ggplot(results, aes(x=Group, y=Burden, color=Group)) + 
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.2),
                     breaks = seq(0, .2, .025)) + ylab("Germline burden") +
  geom_hline(yintercept=0.05, lty=3) + geom_quasirandom(size=0.3) + 
  theme_classic() + theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1))+
  theme(text=element_text(face="bold",size = 16)) 

```

## Session information 

```{r}
sessionInfo()
```